Functional Pearl: the Proof Search Monad

### 2015-10-13 paper/main.tex
Cleanups.

\usepackage{tabularx}
\usepackage{listings}
  \def\li{\lstinline}
  \lstset{
    language=caml,
    flexiblecolumns=false,
    showstringspaces=false,
    basicstyle=\ttfamily,
    framesep=5pt,
    numberstyle=\tiny,
    numbersep=5pt,
    escapeinside={(*}{*)},
    morekeywords={above,abstract,adopts,alias,and,as,assert,begin,below,builtin,consumes,data,do,downto,duplicable,dynamic,else,empty,end,exclusive,explain,fact,fail,flex,for,from,fun,give,if,in,let,match,mutable,open,pack,perm,preserving,rec,take,taking,then,to,type,unknown,val,value,while,with,witness},
    deletekeywords={value},
  }
\usepackage{minted}
  \newminted[ocaml]{ocaml}{mathescape,fontsize=\small}
\usepackage{mathpartir}
  % Improvements to Didier's mathpartir package: make all rule names and
  % references use textsc, and also make references to rules done using the \Rule
  % command clickable in a PDF.
  % \let\TirName\textsc
  % \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
  % \renewcommand{\DefTirName}[1]{\hyperlink{#1}{\TirName {#1}}}
  % \let\DefRule\RefTirName
  % \let\Rule\DefTirName
  \let\Rule\textsc

@@

\begin{abstract}
  We present the proof search monad, a set of combinators that allows one to
  write a proof search engine in a style that resembles closely the inference
  rules. The user calls functions such as \li+premise+, \li+prove+ or
  \li+choice+; the library then takes care of generating a derivation tree.
  Proof search engines written in this style enjoy: first, a one-to-one
  correspondence between the implementation and the theoretical rules, which
  makes manual inspection easier; second, proof witnesses ``for free'', which
  makes a verified, independent validation approach easier too.
\end{abstract}

\section{Theory and practice}

@@

In order to show the validity of this judgement, one will build a proof
derivation using the rules from the logic, shown in \fref{logic} ($[x/y]P$ means
``substitute $x$ with $y$ in $P$''). For instance, proving Equation \ref{ex1}
requires applying \Rule{ExistsE}, then \Rule{Refl}.

@@

For instance, in our logic of existentially-quantified conjunctions of
equalities, one may want to use the algorithmic rules from
\fref{proof-system}. These rules differ from \fref{logic} in that they are
algorithmic; they take an input and return an output.

@@

In particular, in order to determine suitable values for the $x$ parameter in
\Rule{ExistsE}, the implementation reasons in terms of substitutions. $V$ is a
set of variables which may be substituted (recall that free variables are
considered universally quantified, hence not eligible for substitution);
variables that may be substituted are typeset as $\f y$.
The algorithm has internal state, that is, it carries a substitution $\sigma$.
Upon hitting an existential quantifier $\f y$, the algorithmic rules
\emph{open} $\f y$ and mark it as eligible for substitution. Later on (for
instance, upon hitting $\f y = x$), the algorithm may pick a
substitution for $\f y$ using \Rule{Inst}. A substitution may be applied at any
time (\Rule{Subst}). The preconditions of \Rule{Inst} guarantee that the
algorithm makes at most one choice for instantiating $\f y$.

In other words, the algorithmic rules \emph{defer} the \emph{instantiation} of
the existential quantifier until the shape of the proof obligation gives us a
\emph{hint} as to what exactly this instantiation should be. This
\emph{implementation technique} is known as \emph{flexible variables}.

The new algorithmic rules differ from the original logical rules significantly;
first, there are five rules for the algorithmic system, compared to just three
for the logical system. Second, these five rules do not map trivially to their
counterparts in the logical system. Third, these rules are only
algorithmic; the implementation that we are about to roll out uses an
optimized representation for substitutions (union-find), meaning that one not
only needs to check that the algorithmic rules are faithful to the logical
rules, but also that the implementation is faithful to the algorithmic rules.

This paper presents a library that allows one to write an implementation of the
algorithmic rules while automatically generating a derivation. The library
forces the client code to lay out premises, rule applications and
instantiations. The level of detail of the resulting derivation is left up to
the client code; they may wish to record the very compact rules from the logic,
or record more proof steps using the algorithmic rules. In any case, the
derivation serves as a proof witness; if the user wishes to do so, they can
write a validator that takes the witness and verifies that the derivation is,
indeed, correct.

The library has been used, in a preliminary form, to implement the core of the
Mezzo type-checker. This paper presents a cleaned-up, isolated version of this
library.

@@

The logic we present is a much simplified version of the logic (type system)
of Mezzo. In particular, this paper only mentions the right-exists quantifier;
Mezzo has all four possible combinations of left/right exists/forall. Generally,
in proof search, \emph{flexible variables} stem from the right-elimination of
existential quantifiers, or the left-elimination of universal quantifiers. The
right-elimination of universal quantifiers, or the left-elimination of
existential quantifiers give rise to universally-quantified variables, which are
called \emph{rigid}.

@@

Once the algorithm has run, we obtain an output substitution $\sigma$. One can, if they
wish to do so, take the reflexive-transitive closure $\sigma^*$, and apply it to
a flexible variable (say, $\f y$) to recover the parameter of \Rule{ExistsE}
that should be used in the logical rules (here, $x$). This is all very informal
-- the point of the subsequent sections is to formalize the claim that the proof
search algorithm produces a proof witness.

@@

We implement proof search in OCaml (\fref{formulas}); we do not use explicit
substitutions, but rather an optimized representation based on a union-find data
structure. The data type of formulas
is self-explanatory. Variables are implemented as equivalence classes in a
\emph{persistent} union-find data structure, which the module \li+P+ implements.
The $V, \sigma$ parameters in our rules are embodied by the \li+state+ type; just
like the $\sigma$ parameter is chained from one premise to another (\Rule{And}),
\li+state+ is an input and an output to the solver. Just like the $\sigma$ parameter
in the rules, a \li+state+ of the persistent union-find represents
a set of equations between variables. In a sense, \li+state+ is a specific
implementation of the theoretical $\sigma$ parameter. It represents a
substitution; in other words, this is what we want our solver to compute.

The choice of a union-find is irrelevant. All that matters is that we
pick a data structure that models substitutions, and that is \emph{persistent}.
Had we picked an explicit substitution instead of a union-find, the rest of the
discussion would have been the same.

@@

let unify state v1 v2 =                         let rec solve state formula =
  match P.find v1 state, P.find v2 state with     match formula with
  | Flexible, Flexible                            | Equals (v1, v2) ->
  | Flexible, Rigid ->                                unify state v1 v2
      return (P.union v1 v2 state)                | And (f1, f2) ->
  | Rigid, Flexible ->                                solve state f1 >>= fun state ->
      return (P.union v2 v1 state)                    solve state f2
  | Rigid, Rigid ->
      if P.same v1 v2 state then
        return state
      else
        nothing
  \end{ocaml}
  \caption{Solver for the simplified problem}
  \label{fig:solver}
\end{figure}

@@

\section{Building derivations}

@@

We record applications of \Rule{Inst}, \Rule{Refl} and \Rule{And}. This produces
a derivation tree that makes sure that the algorithm follows the algorithmic
rules from \fref{proof-system}. \sref{generate-logical} shows how to generate a
slightly different tree that matches the rules from \fref{logic}.

@@

In the (simplified) sketch from \fref{proof-trees}, rule names are just constant
constructors, since the rule parameters (such as $x$ and $\f y$ in \Rule{Inst})
can be recovered from the \li+formula+. In the general case, the various
constructors of \li+rule_name+ do have parameters that record how one specific
rule was instantiated.

@@

Computations in the writer monad return a result (of type \li+'a+) along with a log of
elements (of type \li+L.a+). The (usual) \li+>>=+ and \li+return+ combinators operate on
the result part of the computation, while the (new) \li+tell+ combinator
operates on the logging part of the computation. This \li+tell+ combinator
appends a new element to the log. Appending elements to the log is done by way of the \li+MONOID+
module type, which essentially demands a value for the empty log, and a function to
append new entries into the log.

In order to get a new \li+>>=+ operator that combines the features of
the option and writer monads, we apply the \li+WriterT+ monad transformer to the
\li+MOption+ monad (\fref{writer}) and obtain \li+MWriter+, a monad whose
computations represent a sequence of derivations (the premises we have proved so far)
along with a result (the \li+state+ that we chain through the premises). These
computations are wrapped in \li+MOption.m+, that is, are wrapped within an
\li+option+ to account for a possible proof failure.

@@

A computation within this new monad has type (simplified after functor applications)
\li+(derivation list * state) option+. It represents a given point in the proof;
the solver is focused on a given rule, has reached a certain state, after
proving a certain list of premises.

We provide a convenience \li+qed+ combinator: once one has obtained the final
state, it pairs the state with the name of the rule we want to conclude with. It
makes the implementation of \li+solve+ (\fref{solver2}) more elegant.

@@

Contrary to the first implementation (\fref{solver}), where the working state
and the return value of \li+solve+ both had type \li+state option+, we now
distinguish between an \li+outcome+ (the result of a call to \li+solve+) and a
working state (a computation in the monad).

An outcome is, as we mentioned earlier, the pair of a final \li+state+ along
with a \li+derivation+ that justifies that we reached this state. The pair is
wrapped in \li+M.t+ (here, \li+option+): proving a formula may fail.

@@

We now have a duality between the \li+outcome+ type (the result of solving a
sub-goal) and the \li+m+ type (a computation within the monad, i.e. a working
state between two premises). Therefore, we introduce two high-level combinators:
\li+premise+ and \li+prove+. The former goes from \li+outcome+ to \li+m+: it
injects a new sub-goal as a premise of the rule we are currently trying to
prove. The latter goes from \li+m+ to \li+outcome+: if all premises have been
satisfied, it draws the horizontal line that builds a new node in the derivation
tree.

@@

  \item \li+fail+ is for situations where no rule applies: this is a failed outcome.
\end{itemize}

@@

\section{Backtracking}

@@

We now extend our formulas with disjunctions (\fref{choice}). A consequence
is that we now need our base monad \li+M+ to offer a new operation; namely, one
that, among several possible choices, picks the first one that is not a failure.
We thus augment \li+MOption+ with a search combinator (\fref{choice}), which in
turn allows one to implement a high-level \li+choice+ combinator for our
library. The \li+choice+ combinator attempts to prove a \li+goal+ by trying a
function \li+f+ on several arguments \li+a+, each of which
is associated to a given rule. We can add one more branch to the \li+solve+
function, which attempts to prove a disjunction by first trying a
left-elimination (\Rule{Or-L}, \fref{proof-system2}), then a right-elimination
(\Rule{Or-R}).

@@

\section{Full example: quantifiers and derivations of logical rules}
\label{sec:generate-logical}

The final section is concerned with adding quantifiers to the language, and
generating derivation trees that reflect the rules of the \emph{logic} rather
than the rules of the \emph{algorithm}. The former is useful for debugging the
inner workings of the algorithm; the latter is useful for producing a proof
certificate.

\begin{figure}
  \centering
\begin{ocaml}
(* Update of the [MyLogic] module. *)
type formula =
  (* ... *)
  | Exists of string * formula
  | Forall of string * formula

@@

(* Update of the [Derivations] module. *)
type derivation =
  goal * rule

@@

(* Update of the [Combinators] module. *)
let prove (goal: Logic.formula) (x: ('a * rule_name) m): 'a outcome =
  M.(x >>= fun (premises, (env, rule)) ->
    return (env, ((env, goal), (rule, Premises premises))))

(* Update of the client. *)
let rec solve (state: state) (goal: formula): state outcome =
  (* ... *)
  | Exists (atom, g) ->
      let var, g, state = open_flexible state atom g in
      let var = assert_open var in
      prove goal begin
        premise (solve state g) >>= fun state ->
        qed (R_ExistsE (name var state)) state
      end
\end{ocaml}
  \caption{Dealing with quantifiers}
  \label{fig:quantifiers}
\end{figure}

We augment the data type of formulas with quantifiers; we replace the type of
rules with the rules from the logic. Furthermore, we demand that the \Rule{Refl}
and \Rule{ExistsE} rules be annotated with their argument.

Bound variables are globally-unique atoms (strings); open variables are
equivalence classes of the union-find, as before.

Now, a \li+goal+ is no longer just a formula: it is a formula interpreted in a
given state. Indeed, since we now open binders and substitute bound variables
for open variables, reading a formula only makes sense in a state where the open
variables of the formula are properly bound in the environment.

We thus change the \li+prove+ combinator to record the state upon creating
a new node in the derivation tree. More precisely, the \li+prove+ combinator
records the state after the premises have been satisfied. This is exactly what
is needed to conclude a rule; for instance, in the \li+Exists+ case, the solver
prods the union-find state to discover the instantiation choice that
\emph{has been made} for the existentially-quantified variable, and records it
in the proof tree.

Here is a pretty-printed version of a derivation tree.

@@

@@

We presented a support library for writing a proof search engine using backtracking in
any given logic; indeed, the library is parameterized by: the type of formulas; the type of
rule applications; the internal state type of the client. By merely using the
combinators of the library, the client gets derivations built for free; this allows
a separate verifier to independently check the steps required to prove the
formula. By opting into the library, the client gets to rewrite their code in a
new syntactic style that makes rule application explicits, forbids ``bundled''
applications of multiple rules at the same time and clearly lays out the
premises required to prove a judgement. Since the code resembles the logical
rules, mistakes are easier to spot.

>>>>

\usepackage{enumerate}
\usepackage{tabularx}
\usepackage{listings}
  \def\li{\lstinline}
  \lstset{
    language=caml,
    flexiblecolumns=false,
    showstringspaces=false,
    basicstyle=\ttfamily,
    framesep=5pt,
    numberstyle=\tiny,
    numbersep=5pt,
    escapeinside={(*}{*)},
    morekeywords={above,abstract,adopts,alias,and,as,assert,begin,below,builtin,consumes,data,do,downto,duplicable,dynamic,else,empty,end,exclusive,explain,fact,fail,flex,for,from,fun,give,if,in,let,match,mutable,open,pack,perm,preserving,rec,take,taking,then,to,type,unknown,val,value,while,with,witness},
    deletekeywords={value},
  }
\usepackage{minted}
  \newminted[ocaml]{ocaml}{mathescape,fontsize=\small}
\usepackage{mathpartir}
  % Improvements to Didier's mathpartir package: make all rule names and
  % references use textsc, and also make references to rules done using the \Rule
  % command clickable in a PDF.
  \let\TirName\textsc
  \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
  \renewcommand{\DefTirName}[1]{\hyperlink{#1}{\TirName {#1}}}
  \let\DefRule\RefTirName
  \let\Rule\DefTirName
  % \let\Rule\textsc

@@

\begin{abstract}
  We present the proof search monad, a set of combinators that allows one to
  write a proof search engine in a style that resembles the formal rules
  closely. The user calls functions such as \li+premise+, \li+prove+ or
  \li+choice+; the library then takes care of generating a derivation tree.
  Proof search engines written in this style enjoy: first, a one-to-one
  correspondence between the implementation and the derivation rules, which
  makes manual inspection easier; second, proof witnesses ``for free'', which
  makes a verified, independent validation approach easier too.
\end{abstract}

\section{Theory and practice}
\label{sec:intro}

@@

In order to show the validity of this judgement, one usually builds a proof
derivation using rules from the logic. In our case, the rules are given in
\fref{logic}; $[x/y]P$ means ``substitute $x$ with $y$ in $P$''. For
instance, proving Equation \ref{ex1} requires applying \Rule{ExistsE}, then
\Rule{Refl}.

@@

For instance, in our logic of existentially-quantified conjunctions of
equalities, one may want to use the rules from \fref{proof-system}. These rules
differ from \fref{logic} in that they are algorithmic; they take an input and
return an output.

@@

In particular, in order to determine suitable values for the $x$ parameter in
\Rule{ExistsE}, the implementation reasons in terms of substitutions. $V$ is a
set of variables which may be substituted (recall that free variables are
considered universally quantified, hence not eligible for substitution);
variables that may be substituted are typeset as $\f y$.
The algorithm has internal state, that is, it carries a substitution $\sigma$.
Upon hitting an existential quantifier $\f y$, the algorithmic rules
\emph{open} $\f y$ and mark it as eligible for substitution (\Rule{ExistsE}).
Later on (for instance, upon hitting $\f y = x$), the algorithm may pick a
substitution for $\f y$ using \Rule{Inst}. A substitution may be applied at any
time (\Rule{Subst}). The preconditions of \Rule{Inst} guarantee that the
algorithm makes at most one choice for instantiating $\f y$.

In other words, the algorithmic rules \emph{defer} the \emph{instantiation} of
the existential quantifier until some sub-goal, later on, gives us a \emph{hint}
as to what exactly this instantiation should be. This \emph{implementation
technique} is known as \emph{flexible variables}.

The new algorithmic rules differ from the original logical rules significantly;
first, there are five rules for the algorithmic system, compared to just three
for the logical system. Second, these five rules do not map trivially to their
counterparts in the logical system. Third, these rules are still very much
abstract; the implementation that we are about to roll out uses an
optimized representation for substitutions (union-find) that is not formalized
in \fref{proof-system}. Phrased differently, one not only needs to check that
the algorithmic rules are faithful to the proof rules, but also that the
implementation itself is faithful to the algorithmic rules.

This paper presents a library that allows one to write an implementation of the
algorithmic rules while automatically generating a derivation. The library
forces the client code to lay out premises, rule applications and
instantiations. The level of detail of the resulting derivation is left up to
the client code; the user may wish to record a proof derivation using the proof
rules, or record a trace of the algorithm using the algorithmic rules.  In any
case, the derivation serves as a witness; in the case of a proof derivation, a
validator may certify that the proof is valid, while in the case of an
algorithmic trace, the user may verify the algorithm, or inspect the trace for
debugging or feedback purposes.

The library has been used, in a preliminary form, to implement the core of the
Mezzo type-checker. This paper presents a cleaned-up, isolated version of this
library that exposes a proper interface using monads and domain-specific
combinators.

@@

The logic we present is a much simplified version of the logic (type system) of
Mezzo. In this document, we only mention the right-exists quantifier. General
systems such as Mezzo have all four possible combinations of left/right
exists/forall.
%
The right-elimination of existential quantifiers, or the left-elimination of
universal quantifiers gives \emph{flexible variables}, while the
right-elimination of universal quantifiers, or the left-elimination of
existential quantifiers gives universally-quantified variables, also called
\emph{rigid variables}.

@@

Once the algorithm has run, we obtain an output substitution $\sigma$. One can, if they
wish to do so, take the reflexive-transitive closure $\sigma^*$, and apply it to
a flexible variable (say, $\f y$) to recover the parameter of \Rule{ExistsE}
that should be used in the logical rules (here, $x$). This way of checking
correctness is not satisfactory and does not scale if nested quantifiers appear
in the goal; the point of the subsequent sections is to make sure the search
algorithm produces a proper proof witness (derivation) that has a tree-like
structure and does not leak implementation details (such as substitutions).

@@

We implement proof search in OCaml (\fref{formulas}); we implement substitutions
using a union-find data structure. The data type of formulas
is self-explanatory. Variables are implemented as equivalence classes in the
\emph{persistent} union-find data structure, which the module \li+P+ implements.
The $V, \sigma$ parameters in our rules are embodied by the \li+state+ type; just
like the $\sigma$ parameter is chained from one premise to another (\Rule{And}),
\li+state+ is an input and an output to the solver. Just like the $\sigma$ parameter
in the rules, a \li+state+ of the persistent union-find represents
a set of equations between variables. The algorithmic rules mentioned a
theoretical $\sigma$ parameter; the \li+state+ is our specific implementation
choice.

The choice of a union-find (as opposed to explicit substitutions) is irrelevant.
All that matters is that we pick a data structure that models substitutions, and
that the structure be \emph{persistent}.

@@

let unify state v1 v2: state =                  let rec solve state formula: state =
  match P.find v1 state, P.find v2 state with     match formula with
  | Flexible, Flexible                            | Equals (v1, v2) ->
  | Flexible, Rigid ->                                unify state v1 v2
      return (P.union v1 v2 state)                | And (f1, f2) ->
  | Rigid, Flexible ->                                solve state f1 >>= fun state ->
      return (P.union v2 v1 state)                    solve state f2
  | Rigid, Rigid ->
      if P.same v1 v2 state then
        return state
      else
        nothing
  \end{ocaml}
  \caption{Solver for the simplified problem}
  \label{fig:solver}
\end{figure}

@@

\section{Building derivations}
\label{sec:derivations}

@@

We record applications of \Rule{Inst}, \Rule{Refl} and \Rule{And}. This produces
a derivation tree (algorithm trace) that makes sure that the algorithm follows
the algorithmic rules from \fref{proof-system}. \sref{generate-logical} shows
how to generate a different, more compact tree that matches the rules from
\fref{logic}.

@@

In the (simplified) sketch from \fref{proof-trees}, rule names are just constant
constructors, since the rule parameters (such as $x$ and $\f y$ in \Rule{Inst})
can be recovered from the \li+formula+. In the general case
(\sref{generate-logical}), the various constructors of \li+rule_name+ do have
parameters that record how one specific rule was instantiated.

@@

Computations in the writer monad return a result (of type \li+'a+) along with a log of
elements (of type \li+L.a+). The (usual) \li+>>=+ and \li+return+ combinators operate on
the result part of the computation, while the (new) \li+tell+ combinator
operates on the logging part of the computation. The \li+tell+ combinator
appends a new element to the log; this is done by way of the \li+MONOID+ module
type, which essentially demands a value for the \li+empty+ log, and a function
to \li+append+ new entries into the log.

In order to get a new \li+>>=+ operator that combines the features of
the option and writer monads, we apply the \li+WriterT+ monad transformer to the
\li+MOption+ monad (\fref{writer}) and obtain \li+MWriter+.

@@

The type of computations \li+'a MWriter.t+ boils down (after functor
application) to \li+(derivation list * state) option+. A computation in the monad
represents a given point in the proof; the solver is focused on a rule; has
proved a number of premises so far (the \li+derivation list+); has reached a
certain \li+state+ (threaded through the premises). The \li+option+ type
accounts for failure; in case a premise cannot be proved, the computation aborts
and becomes \li+None+.

@@

Contrary to the first implementation (\fref{solver}), where the working state
and the return value of \li+solve+ both had type \li+state option+, we now
distinguish between an \li+outcome+ (the result of a call to \li+solve+) and a
working state (a computation in the \li+MWriter+ monad).

An \li+outcome+ is the pair of a final \li+state+ along
with a \li+derivation+ that justifies that we reached this state. The pair is
wrapped in \li+M.t+ (here, \li+option+): if the computation of premises is a
failure, then the proof of the desired goal is a failure too.

@@

We now have a duality between the \li+outcome+ type (the result of solving a
goal) and the \li+m+ type (a computation within the monad, i.e. a working
state between two premises). Therefore, we introduce two high-level combinators:
\li+premise+ and \li+prove+. The former goes from \li+outcome+ to \li+m+: it
injects a new sub-goal as a premise of the rule we are currently trying to
prove. The latter goes from \li+m+ to \li+outcome+: if all premises have been
satisfied, it draws the horizontal line that builds a new node in the derivation
tree.

@@

  \item \li+fail+ is for situations where no rule applies: this is a failed outcome.

  \item \li+qed+ is a convenience combinator that pairs the state with the name
    of the rule we want to conclude with; it makes the implementation of
    \li+solve+ (\fref{solver2}) more elegant.

\end{itemize}

@@

\section{Backtracking}
\label{sec:backtracking}

@@

We now extend our formulas with disjunctions (\fref{choice}). A consequence
is that we now need our base monad \li+M+ to offer a new operation; namely, one
that, among several possible choices, picks the first one that is not a failure.
We thus augment \li+MOption+ with a search combinator (\fref{choice}), which in
turn allows us to implement a high-level \li+choice+ combinator for our
library. The \li+choice+ combinator attempts to prove a \li+goal+ by trying a
function \li+f+ on several arguments of type \li+a+, each of which
has a given \li+outcome+. We extend \li+solve+ with an extra case,
which attempts to prove a disjunction by first trying a
left-elimination (\Rule{Or-L}, \fref{proof-system2}), then a right-elimination
(\Rule{Or-R}).

@@

\section{Extension: quantifiers and proof trees}
\label{sec:generate-logical}

We mentioned earlier that the derivation we were building tracked the
application of algorithmic rules; that is, we were building a \emph{trace} of
the algorithm. While the trace is useful to extract information for the user,
one may also want to build a proper proof witness in order to certify the
validity of the formula.

In order to make a proof tree relevant and not just provide the substitution as
the proof witness, we introduce quantifiers to the language, and construct proof
trees that apply the proof rules from \fref{logic}, \fref{proof-system2},
\fref{logic-quantifiers}. The nodes of the proof tree are rules; each node is
annotated, if applicable, by its implicit parameters. That is, \Rule{Refl} and
\Rule{ExistsE} are annotated with their implicit $x$ parameter.

\begin{figure}
  \centering
  \begin{mathpar}
    \inferrule[ForallE]{
      P
    }{
      \forall y.\ P
    }
  \end{mathpar}
  \caption{Extra rule for the universal quantifier}
  \label{fig:logic-quantifiers}
\end{figure}

\begin{figure}
  \centering
\begin{ocaml}
(* i) Update of the [MyLogic] module. *)
type formula =
  (* ... *)
  | Exists of string * formula
  | Forall of string * formula

@@

(* ii) Update of the [Derivations] module. *)
type derivation =
  goal * rule

@@

(* iii) Update of the [Combinators] module. *)
let prove (goal: Logic.formula) (x: ('a * rule_name) m): 'a outcome =
  M.(x >>= fun (premises, (env, rule)) ->
    return (env, ((env, goal), (rule, Premises premises))))

(* iv) Update of the client. *)
let rec solve (state: state) (goal: formula): state outcome =
  (* ... *)
  | Exists (atom, g) ->
      let var, g, state = open_flexible state atom g in
      let var = assert_open var in
      prove goal begin
        premise (solve state g) >>= fun state ->
        qed (R_ExistsE (name var state)) state
      end
\end{ocaml}
  \caption{Dealing with quantifiers}
  \label{fig:quantifiers}
\end{figure}

The updates to the library required to implement quantifiers are minimal; the
bulk of the work is essentially writing substitution and a proper treatment of
binders on the client-side.

\fref{quantifiers} presents in an informal style the series of updates
required.
\begin{enumerate}[i)]
  \item We augment the data type of formulas with quantifiers; we replace the
    type of rules with the rules from the logic. Furthermore, we demand that the
    \Rule{Refl} and \Rule{ExistsE} rules be annotated with their argument.

    Bound variables are globally-unique atoms (strings); open variables are
    equivalence classes of the union-find, as before (not shown here).

  \item We previously did not distinguish between a \li+goal+ and a
    \li+formula+; this was only possible because we assumed all variables were
    initially open, meaning that we could deference an open variable in any
    \li+state+. Now, we open binders and substitute variables, through the
    allocation of new points in the union-find (the \li+state+). Therefore, a
    given \li+formula+ only makes sense when paired with a specific \li+state+.

  \item We update the \li+prove+ combinator to record the \li+state+ upon
    creating a new node in the derivation tree. More precisely, the \li+prove+
    combinator records the \li+state+ after the premises have been satisfied.

  \item Only a slight is needed on the client side to record a proper proof
    witness: in the \li+Exists+ case, the solver prods the union-find state to
    discover the instantiation choice that \emph{has been made} for the
    existentially-quantified variable, and records it in the proof tree.
\end{enumerate}

The sample code in the library comes with a pretty-printer. Here is the output
for a simple formula that combines all features from our formula language.

@@

\section{Source code}

The library is available online at
\url{https://github.com/msprotz/proof-search-monad}. The file \li+example01.ml+
contains the full implementation of the primitive solver described in
\sref{intro}. The file \li+example02.ml+ contains the backtracking solver
written within the proof search monad, as described in \sref{backtracking}. One
can get the non-backtracking version, described in \sref{derivations}, by
replacing \li+MExplore+ with \li+MOption+. Finally, the file \li+example03.ml+
contains the final algorithm described in \sref{generate-logical}. The
representation of binders adopted in the last example is suboptimal; bound
variables are represented using globally-unique atoms. One may want to use
locally nameless, with De Bruijn indices for bound variables (as used in Mezzo).
It allows keeps the boilerplate to a minimum, though.

\section{Related work}

ZOMG TODO

@@

We presented a support library for writing a proof search engine using
backtracking. The library is parameterized by: the type of formulas; the type of
rule applications; the internal state type of the client. This leaves complete
freedom for the client to define their own logic. By merely using the
combinators of the library, the client gets derivations built for free; this allows
a separate verifier to independently check the steps required to prove the
formula. By opting into the library, the client gets to rewrite their code in a
new syntactic style that makes rule application explicits, forbids ``bundled''
applications of multiple rules at the same time and clearly lays out the
premises required to prove a judgement. Since the code resembles the logical
rules, mistakes are easier to spot.
